Parsing Postfix Log Files

### 2009-04-17 doc/logparser-background.tex
Some minor changes.
Corrections based on proofreading to the preface and chapters 1-3.

A short example of the optimisation possible using data from the database
is determining which Postfix restrictions reject the highest number of
mails:

@@

If the database supports sub-selects (where the results of one query are
used as a parameter in another), percentages can be obtained for the top
ten restrictions (sample output shown in \tableref{Sample output from SQL
query}):

\input{build/include-sample-sql-query}

\begin{table}[thbp]
    \caption{Sample output from SQL query, showing the percentage of
    delivery attempts rejected by top ten restrictions}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}

    All percentages above are exclusive, i.e.\ the first restriction's
    number of rejections does not include the second restriction's number
    of rejections.

@@

Another example is determining which restrictions are not effective: this
sample query shows which restrictions had fewer than 100 rejections in the
last log file parsed, and the percentage of total rejections each of those
restrictions represents.

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE action = 'DELIVERY_REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE action = 'DELIVERY_REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

The database queries above yield summary statistics about the efficiency of
anti-spam techniques.  These statistics are far less feasible to assess
directly from log files without prior pre-processing into a database in the
fashion proposed, implemented, and tested herein.

@@

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: Alice in Distribution alice@example.com
501 5.1.7 Bad sender address syntax
RCPT TO: Bob in Sales/Marketing bob@example.com
503 5.5.1 Error: need MAIL command
DATA
503 5.5.1 Error: need RCPT command
Message headers and body sent here.
.
502 5.5.2 Error: command not recognized
QUIT
221 2.0.0 Bye
\end{verbatim}

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately, other programs,
particularly newer programs, were written by competent programmers or send
mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
Microsoft Outlook on Windows hosts).  Traditionally a mail server would
have done its best to deal with deficient clients, with the intention of
accepting as much mail sent to its users as possible, e.g.\ by accepting
sender or recipient addresses that were not enclosed in \texttt{<>}, or by
ignoring the absence of a HELO command.  The HELO command is the first
command sent by the client in the \acronym{SMTP} conversation; it is
required by the protocol, but its original purpose was to detect a server
trying to deliver mail to itself, so some implementations omitted it.

@@

Postfix is a \glsfirst{MTA} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive optional anti-spam restrictions,
allowing an administrator to employ those restrictions which they judge
suitable for their server's needs, rather than a fixed set chosen by
Postfix's author.  These restrictions can be selectively applied, combined,
and bypassed on a per-client, per-recipient, or per-sender basis, allowing
different levels of stricture and permissiveness.  Administrators can
supply their own rejection messages to make it clear to senders exactly why
their mail was rejected.  Policy servers (\sectionref{policy servers})
provide a simple way to write new restrictions without having to modify
Postfix's source code.  Unfortunately this flexibility has a cost:
complexity in the log files generated.  Although it is relatively simple in
most cases to use standard Unix text processing utilities to determine the
fate of an individual mail, sometimes it can be quite difficult.  For
most mails the journey is simple and brief, but the remainder can have
quite complex journeys (see \sectionref{complications} for details).

@@

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}; the following is only a brief
introduction.

@@

\end{eqlist}

The restriction \texttt{check\_client\_access} checks if the \acronym{IP}
address of the connected client is found in the specified table and returns
the associated action if found; the method of searching the file is
dependant on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\newline{}
\tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
\acronym{DNSBL} \texttt{dnsbl.example.com} and rejects the command if the
client's \acronym{IP} address is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno}, or the name of another restriction to be
evaluated; other less commonly used results are possible as described
in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.  The
meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing
exceptions to more general rules.  E.g.\ the general rule is that machines
on the local network must authenticate before sending mail, except for the
web server, because the legacy applications running on it lack
authentication support; this configuration example shows how:

@@

That example also shows that it is trivial to supply a custom rejection
message.  When the result is the name of another restriction Postfix will
evaluate that restriction, allowing restrictions to be chosen based on the
client \acronym{IP} address, client hostname, HELO hostname, sender
address, or recipient address.  E.g.\ the administrator may require that
all clients on the local network have valid DNS entries, to prevent people
sending mail from unknown machines; one example of how to achieve this is:

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  E.g.\ a phishing mail might
claim that the payroll system had a disastrous disk failure, and until the
server is replaced all salary payments will have to be processed manually,
so please reply to this mail with your name, address, and bank account
details; the criminal can then use any details sent to him to help with
identity theft.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote
        \acronym{IP} address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that failed
        temporarily.  Sadly spam senders are using increasingly complex and
        well written programs to distribute spam, frequently using an ISP
        provided \acronym{SMTP} server from a compromised machine on the
        ISP's network.  Greylisting will slowly become less effective as
        spam senders adapt, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used when generating the results in
        \sectionref{Evaluation} are shown in \tableref{Most effective
        Postfix restrictions}.  The table shows that greylisting is worth
        using at the moment, particularly when you take into account its
        position as the final restriction that a mail must overcome in the
        configuration used on the mail server that generated the log files:
        on that server greylisting only takes effect for mails that have
        passed all other restrictions.  Some problems may be encountered
        when using greylisting: some servers fail to retry after a
        temporary failure, or legitimate mail may be delayed, particularly
        when coming from a pool of servers.

@@

    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
        perform tests on features of the delivery attempt (e.g.\
        \acronym{IP} address, sender address), incrementing or decrementing
        a score based on the results; if the eventual score is higher than
        a threshold the mail is rejected.  The administrator must manually
        whitelist clients if they are to bypass a Postfix restriction,
        whereas using a threshold that requires a delivery attempt to hit
        several scored restrictions will allow mail from clients that would
        fall foul of one binary restriction only.

>>>>

Determining which Postfix restrictions reject the highest number of
delivery attempts is a short example of the optimisation possible using
data from the database:

@@

If the database supports sub-selects (where the results of one query are
used as a parameter in another), percentages can be obtained for the top
ten restrictions using the query in \figureref{Sample SQL query}, producing
output similar to \tableref{Sample output from SQL query}.

\begin{figure}[thbp]
    \caption{Sample SQL query showing the percentage of delivery attempts
    rejected by the top ten restrictions}
    \empty{}\label{Sample SQL query}
    \input{build/include-sample-sql-query}
\end{figure}

\begin{table}[thbp]
    \caption{Sample output from an SQL query showing the percentage of
    delivery attempts rejected by the top ten restrictions}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}

    All percentages in \tableref{Sample output from SQL query} are
    exclusive, i.e.\ the first restriction's number of rejections does not
    include the second restriction's number of rejections.

@@

Another example is determining which restrictions are not effective:
\figureref{Sample SQL query showing ineffective restrictions} shows which
restrictions rejected fewer than 100 delivery attempts in the last log file
parsed, and the percentage of total rejections in that log file that each
of those restrictions represents.

\begin{figure}[thbp]
    \caption{Sample SQL query showing ineffective restrictions}
    \empty{}\label{Sample SQL query showing ineffective restrictions}
    \begin{verbatim}
    SELECT name, description, restriction_name, hits,
            (hits * 100.0 /
                (SELECT SUM(hits)
                    FROM rules
                    WHERE action = 'DELIVERY_REJECTED'
                )
            ) || '%' AS percentage
        FROM rules
        WHERE action = 'DELIVERY_REJECTED'
            AND hits < 100
        ORDER BY hits ASC;
    \end{verbatim}
\end{figure}

The sample database queries yield summary statistics about the efficiency of
anti-spam techniques.  These statistics are far less feasible to assess
directly from log files without prior pre-processing into a database in the
fashion proposed, implemented, and tested herein.

@@

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: Alice in Distribution alice@example.com
501 5.1.7 Bad sender address syntax
RCPT TO: Bob in Sales/Marketing bob@example.com
503 5.5.1 Error: need MAIL command
DATA
503 5.5.1 Error: need RCPT command
Message headers and body sent here.
502 5.5.2 Error: command not recognized
\end{verbatim}

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately, newer programs
tend to be written by competent programmers or send mail using well written
programs (e.g.\ Postfix or Sendmail on Unix hosts, Microsoft Outlook on
Windows hosts).  Traditionally a mail server would have done its best to
deal with deficient clients, with the intention of accepting as much mail
sent to its users as possible, e.g.\ by accepting sender or recipient
addresses that were not enclosed in \texttt{<>}, or by ignoring the absence
of a HELO command.\footnote{The HELO command is the first command sent by
the client in the \acronym{SMTP} conversation; it is required by the
protocol, but its original purpose was to detect a server trying to deliver
mail to itself, so some implementations omitted it.}  Given the volume of
spam sent daily, this liberal approach is no longer viable.

@@

Postfix is a \glsfirst{MTA} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive optional anti-spam restrictions,
allowing an administrator to employ those restrictions which they judge
suitable for their server's needs, rather than a fixed set chosen by
Postfix's author.  These restrictions can be selectively applied, combined,
and bypassed on a per-client, per-recipient, or per-sender basis, allowing
different levels of stricture and permissiveness.  Administrators can
supply their own rejection messages to make it clear to senders why their
mail was rejected.  Policy servers (\sectionref{policy servers}) provide a
simple way to write new restrictions without having to modify Postfix's
source code.  Unfortunately this flexibility has a cost: the complexity
Postfix's log files.  Although it is usually relatively simple to use
standard Unix text processing utilities to determine the fate of an
individual mail, sometimes it can be quite difficult.  For most mails the
journey is simple and brief, but the remainder can be quite complex (see
\sectionref{complications} for details).

@@

Postfix restrictions are documented fully in~\cite{smtpd_per_user_control,
policy-servers, smtpd_access_readme}; the following is only a brief
introduction.

@@

\end{boldeqlist}

The restriction \texttt{check\_client\_access} checks if the IP address of
the connected client is found in the specified table and returns the
associated action if found; the method of searching the file is dependant
on the type of the lookup table~\cite{postfix-lookup-tables}.  Other
restrictions determine their result by consulting external sources,
e.g.\newline{} \tab{}\texttt{reject\_rbl\_client
dnsbl.example.com}\newline{} checks the \acronym{DNSBL}
\texttt{dnsbl.example.com} and rejects the command if the client's IP
address is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno}, or the name of another restriction to be
evaluated; other less commonly used results are possible as described
in~\cite{access-man-page}.  The meaning of \textit{permit\/} and
\textit{reject\/} is obvious; \textit{dunno\/} means to stop evaluating the
current restriction and continue processing the remainder of the
restriction list, allowing exceptions to more general rules.  The
configuration example below shows how to require that all machines on the
local network except for the web server authenticate before sending mail;
the web server is exempt because the legacy applications running on it lack
authentication support.  The restriction list is evaluated from top to
bottom: \texttt{permit\_sasl\_authenticated} permits authenticated clients;
an exception is made for the web server in the
\texttt{check\_client\_access} check, and other local machines are
rejected, because they must be unauthenticated to reach that point in the
restriction list.

@@

That example also shows that it is trivial to supply a custom rejection
message.  When the result is the name of another restriction Postfix will
evaluate that restriction, allowing restrictions to be chosen based on the
client IP address, client hostname, HELO hostname, sender address, or
recipient address.  E.g.\ the administrator may require that all clients on
the local network have valid DNS entries, to prevent people sending mail
from unknown machines; one example of how to achieve this is:

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  E.g.\ a phishing mail might
claim that the payroll system had a disastrous disk failure, and until the
server is replaced all salary payments will have to be processed manually,
so please reply to this mail with your name, address, and bank account
details; the criminal can then use any details sent to him to help with
identity theft.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote IP
        address)}\newline{} has not been seen before; on second and
        subsequent delivery attempts from that triple the mail will be
        accepted.  This blocks spam from some senders because maintaining a
        list of failed addresses and retrying after a temporary failure is
        uneconomical for a spam sender, but a legitimate mail server must
        retry deliveries that fail temporarily.  Sadly spam senders are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided \acronym{SMTP} server from a
        compromised machine on the ISP's network.  Greylisting will slowly
        become less effective as spam senders adapt, but it does block a
        large percentage of spam mail at the moment; the most effective
        restrictions from the \numberOFlogFILES{} log files used when
        generating the results in \sectionref{Evaluation} are shown in
        \tableref{Most effective Postfix restrictions}.  The table shows
        that greylisting is worth using at the moment, particularly when
        you take into account its position as the final restriction that a
        mail must overcome in the configuration used on the mail server
        that generated the log files: on that server greylisting only takes
        effect for mails that have passed all other restrictions.  Some
        problems may be encountered when using greylisting: some servers
        fail to retry after a temporary failure, or legitimate mail may be
        delayed, particularly when coming from a pool of servers.

@@

    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
        perform tests on features of the delivery attempt (e.g.\ IP
        address, sender address), incrementing or decrementing a score
        based on the results; if the final score is higher than a threshold
        the delivery attempt is rejected.  The administrator must manually
        whitelist clients if they are to bypass a Postfix restriction,
        whereas using a threshold that requires a delivery attempt to hit
        several scored restrictions will allow mail from clients that would
        be rejected by a binary restriction.

### 2009-04-17 doc/logparser-complications.tex
Corrections based on proofreading to the preface and chapters 1-3.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  The crux of this complication is
that re-injected mails appear in the log files without explicit logging
indicating their source.

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \acronym{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  These heuristics are quite
restrictive, and it appears to have little scope for false positives; any
false positives will cause a warning when the next log line for that mail
is parsed.  False negatives are less likely to be detected: there may be
queueid clashes (and warnings) if mails remain in the state tables after
they should have been removed, otherwise the only way to detect false
negatives is to examine the state tables after each parsing run.

@@

    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
        that passes normally through Postfix will have a \daemon{cleanup}
        line; lack of one is a sure sign the mail did not make it too far.

>>>>

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  The crux of this complication is
that re-injected mails appear in the log files without explicit logging
indicating their source.  Unfortunately, it is not possible to distinguish
between child mails waiting to be tracked and other mails whose origin is
unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} can not be expected to fully understand and reconstruct a
mail when it is lacking some of that mail's log lines.

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).  These heuristics are quite restrictive,
and it appears to have little scope for false positives; any false
positives will cause a warning when the next log line for that mail is
parsed.  False negatives are less likely to be detected: there may be
queueid clashes (and warnings) if mails remain in the state tables after
they should have been removed, otherwise the only way to detect false
negatives is to examine the state tables after each parsing run.

@@

    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
        that passes normally through Postfix will have a \daemon{cleanup}
        log line; lack of one is a sure sign the mail did not make it too
        far.

### 2009-04-17 doc/logparser-implementation.tex
Some minor changes.
Corrections based on proofreading to the preface and chapters 1-3.

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected an \acronym{SMTP} command;
the individual fields will be described in \sectionref{connections table}.
The results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the Database Schema picture}, to complement the
in-depth descriptions of each table.

@@

\newlength{\belowcaptionskipORIG}
\setlength{\belowcaptionskipORIG}{\belowcaptionskip}
\setlength{\belowcaptionskip}{10pt}
\showgraph{build/database-schema}{Diagram of the Database Schema}{Diagram
of the Database Schema picture}
\setlength{\belowcaptionskip}{\belowcaptionskipORIG}

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line, this field should describe the event causing the log
        lines this rule recognises.

@@

    \item [server\_ip] The \acronym{IP} address of the server: the local
        address when receiving mail, the remote address when sending mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \acronym{IP} address could not be resolved
        to a hostname via DNS\@.

    \item [client\_ip] The client \acronym{IP} address: the remote address
        when receiving mail, the local address when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \acronym{IP} address could not be resolved
        to a hostname via DNS\@.

@@

    \item How long it takes to parse each log file (\graphref{parsing time
        vs file size vs number of log lines graph}).

    \item The number of rules tried when recognising log lines
        (\graphref{Mean number of rules tried per log line}); the framework
        may need to try multiple rules for a given log line before it finds
        one that recognises it.

    \item The number of log lines and rules tried for each Postfix
        component (\graphref{Mean number of rules tried per log line for
        each Postfix component}).

@@

    \item [Pre-process the file.]  The new algorithm leverages the
        knowledge gained when writing rules and developing \parsername{};
        it performs many substitutions on the input log lines, replacing
        commonly occurring variable terms (e.g.\ email addresses,
        \acronym{IP} addresses, the standard start of rejection messages)
        with keywords described in \sectionref{regex components}.  The
        purpose of this step is to utilise existing knowledge to create
        more accurate regexes; it replaces many variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The altered log lines are written to a temporary file,
        which subsequent stages will use instead of the original input
        file.

>>>>

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection that rejected an \acronym{SMTP} command;
the individual fields will be described in \sectionref{connections table}.
The results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the database schema picture}, to complement the
in-depth descriptions of each table.

@@

\newlength{\belowcaptionskipORIG}
\setlength{\belowcaptionskipORIG}{\belowcaptionskip}
\setlength{\belowcaptionskip}{10pt}
\showgraph{build/database-schema}{Diagram of the database schema}{Diagram
of the database schema picture}
\setlength{\belowcaptionskip}{\belowcaptionskipORIG}

@@

    \item [description] Something must have occurred to cause Postfix to
        output each log line; this field should describe the event causing
        the log lines this rule recognises.

@@

    \item [server\_ip] The IP address of the server: the local address when
        receiving mail, the remote address when sending mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

    \item [client\_ip] The client IP address: the remote address when
        receiving mail, the local address when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item How long it takes to parse each log file (\graphref{parsing time
        vs log file size vs number of log lines graph}).

    \item The number of rules used when recognising log lines
        (\graphref{Mean number of rules used per log line}); the framework
        may need to try multiple rules for a given log line before it finds
        one that recognises it.

    \item The number of log lines and rules used for each Postfix component
        (\graphref{Mean number of rules used per log line for each Postfix
        component}).

@@

    \item [Pre-process the file.]  The new algorithm leverages the
        knowledge gained when writing rules and developing \parsername{};
        it performs many substitutions on the input log lines, replacing
        commonly occurring variable terms (e.g.\ email addresses, IP
        addresses, the standard start of rejection messages) with keywords
        described in \sectionref{regex components}.  The purpose of this
        step is to utilise existing knowledge to create more accurate
        regexes; it replaces many variable words with fixed words,
        improving the subsequent classification of words as fixed or
        variable.  The altered log lines are written to a temporary file,
        which subsequent stages will use instead of the original input
        file.

### 2009-04-17 doc/logparser-introduction.tex
Corrections based on proofreading to the preface and chapters 1-3.

The approach chosen to measure performance is to analyse the log files
produced by the \acronym{SMTP} server in use, Postfix, rather than
modifying its source code to generate statistics: this approach improves
the chances of other Postfix users testing and using the software developed
for this project.  The need arose for a parser capable of dealing with the
great number and variety of log lines produced by Postfix: the parser must
be designed and implemented so that adding support for parsing new inputs
is a simple task, because the log lines to be parsed will change over time.
This variety in log lines occurs for several reasons:

@@

    \item Policy servers~\cite{policy-servers} may log different messages
        depending on the characteristics of the connection.

@@

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam techniques are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2000 computers and over 1800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of log lines is not fixed, and log
lines can differ in many ways between servers, even within the same
organisation, where servers may be configured differently or running
different versions of Postfix.

It was hoped to reuse an existing parser rather than writing one from
scratch, but the effort required to adapt and improve an existing parser
was judged to be greater than the effort to write a new one, because the
techniques used by the existing parsers severely limited their potential:

    \item Some parsers ignored the majority of log lines, parsing specific
        log lines accurately, but without any provision for parsing new or
        similar log lines.

    \item Other parsers sloppily parsed the majority of log lines, but were
        incapable of distinguishing between log lines in one same category,
        e.g.\ not distinguishing between different anti-spam techniques
        causing the rejection of delivery attempts.

    \item None of the reviewed parsers extracted enough data for future
        in-depth analysis.

The only prior published work on the subject of parsing Postfix log files
that the author is aware of is \textit{Log Mail Analyser: Architecture and
Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
data from log files, correlate it, and present it in a form suitable for a
systems administrator to search using the myriad of standard Unix text
processing utilities already available; it is reviewed alongside the other
parsers in the State of the Art Review in chapter~\ref{state of the art
review}.

Once it was decided that a new parser would be written, an architecture was
required to base the implementation on.  Existing architectures are
tailored towards parsing inputs with a fixed grammar or a tightly
constrained format, whereas Postfix log files lack any form of constraint,
as outlined earlier.  A new architecture was designed and developed for
this parser, with the hope that it will be useful to others.  The resulting
architecture is conceptually simple: provide a few generic functions
(\textit{actions\/}), each capable of dealing with an entire category of
inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a
multitude of precise patterns (\textit{rules\/}), each of which recognises
one input variant and only that variant (e.g.\ rejection by a specific
\acronym{DNSBL}), and specifies which action will process the inputs it
recognises.  This architecture is ideally suited to parsing inputs
where the input is not fully understood or does not conform to a fixed
grammar: the architecture warns about unparsed inputs and other errors, but
continues parsing as best it can, allowing the developer of a new parser to
decide which deficiencies are most important and require immediate
attention, rather than being forced to fix the first error that arises.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring much
work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam defences, to produce statistics
showing how effective each technique in use is, or to provide a baseline to
test new anti-spam measures against.  Numerous other uses are possible for
such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times, or providing a dedicated service for them; detecting virus
outbreaks that propagate via mail; as a base for billing customers on a
shared server.  Preserving the raw data enables users to develop a
multitude of uses far beyond those conceived of by the author.

@@

Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, discussing
the various optimisations implemented in the framework, and the effect they
have.  It also discusses the coverage achieved by \parsername{} over
\numberOFlogFILES{} log files, with separate sections for the number of log
lines correctly recognised, and the number of connections and mails whose
journey through Postfix was correctly reconstructed.

@@

Appendix~\ref{Postfix Daemons} concludes this thesis with a brief
description of Postfix daemons.

>>>>

The approach chosen to measure performance is to analyse the log files
produced by the \acronym{MTA} in use, Postfix, rather than modifying its
source code to generate statistics: this approach improves the chances of
other Postfix users testing and using the software developed for this
project.  The need arose for a parser capable of dealing with the great
number and variety of log lines produced by Postfix: the parser must be
designed and implemented so that adding support for parsing new inputs is a
simple task, because the log lines to be parsed will change over time.
This variety in log lines occurs for several reasons:

@@

    \item Policy servers (see \sectionref{policy servers}) may log
        different messages depending on the characteristics of the
        connection.

@@

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam techniques are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2000 computers and over 1800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis when a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of log lines is not fixed, and log
lines can differ in many ways between servers, even within the same
organisation, where servers may be configured differently or running
different versions of Postfix.

The only prior published work on the subject of parsing Postfix log files
that the author is aware of is \textit{Log Mail Analyser: Architecture and
Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
data from Postfix log files, correlate it, and present it in a form
suitable for a systems administrator to search using the myriad of standard
Unix text processing utilities already available; it is reviewed alongside
the other parsers in the State of the Art Review in chapter~\ref{state of
the art review}.  It was hoped to reuse an existing parser rather than
writing one from scratch, but the effort required to adapt and improve an
existing parser was judged to be greater than the effort to write a new
one, as described in the State of the Art Review.

Once it was decided that a new parser would be written, an architecture was
required to base the implementation on.  Existing architectures are
tailored towards parsing inputs with a fixed grammar or a tightly
constrained format, whereas Postfix log files lack any form of constraint,
as outlined earlier.  A new architecture was designed and developed for
this parser, with the hope that it will be useful to others.  The resulting
architecture is conceptually simple: provide a few generic functions
(\textit{actions\/}), each capable of dealing with an entire category of
inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a
multitude of precise patterns (\textit{rules\/}), each of which recognises
one input variant and only that variant (e.g.\ rejection by a specific
\acronym{DNSBL}), and specifies which action will process the inputs it
recognises.  This architecture is ideally suited to parsing inputs that are
not fully understood or do not conform to a fixed grammar: the architecture
warns about unrecognised inputs and other errors, but continues parsing as
best it can, allowing the developer of a new parser to decide which
deficiencies are most important and require immediate attention, rather
than being forced to fix the first error that arises.
The architecture is designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring much
work or a high level of understanding of the parsing process and the
parser's internal workings.

This architecture is the basis of \parsername{}, a program that parses
Postfix log files and places the resulting data into a database for later
analysis.  The gathered data can be used to optimise current anti-spam
defences, to produce statistics showing how effective each technique in use
is, or to provide a baseline to test new anti-spam measures against.
Numerous other uses are possible for such data: improving server
performance by identifying troublesome destinations and reconfiguring
appropriately; identifying regular high volume uses (e.g.\ customer
newsletters) and restricting those uses to off-peak times, or providing a
dedicated service for them; detecting virus outbreaks that propagate via
mail; as a base for billing customers on a shared server.  Preserving the
raw data enables a multitude of uses far beyond those conceived of by the
author.

@@

Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, discussing
the various optimisations implemented in the parser, and the effect they
have.  It also discusses the coverage achieved by \parsername{} over
\numberOFlogFILES{} log files, with separate sections for the number of log
lines correctly recognised, and the number of connections and mails whose
journey through Postfix was correctly reconstructed.

@@

Appendix~\ref{Postfix Daemons} provides a brief description of Postfix
daemons.
